---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- read10xCounts(samples = c("pbmc5k"="/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix"), col.names = TRUE)
  
sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
sce <- sce[, colSums(assay(sce, "counts")) > 0]
sce

```

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(sce,
                                assay.type = "counts",
                                lower = 100,
                                niters = 10000)
out_df <- as.data.frame(out)
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}

positions <- which(out$FDR < 0.01)

sce_filtered_emptydrops <- sce[, positions]
sce_filtered_emptydrops

remove(positions)
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
plot_data <- tibble(library_size = colSums(assay(sce, "counts")), droplet_status = (out$FDR < 0.01))

plot_data$droplet_status[is.na(plot_data$droplet_status)] <- "empty"
  
plot_data$droplet_status <- factor(as.character(plot_data$droplet_status), levels = c("empty", "FALSE", "TRUE"), labels = c("not_tested", "empty droplet", "cell"))
  
ggplot(plot_data, aes(x = droplet_status, y = library_size)) + geom_violin() + scale_y_log10()
  
  
unique(plot_data$droplet_status)
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 


all_barcodes <- colnames(sce)

compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_emptydrops)

colData(sce_scdblfinder)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}
sce_filtered_scdblfinder <- sce_scdblfinder[,sce_scdblfinder$scDblFinder.class == "singlet"]
sce_filtered_scdblfinder
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
plot_data <- tibble(library_size = colSums(assay(sce_scdblfinder, "counts")), scDblFinder_status = sce_scdblfinder$scDblFinder.class)

ggplot(plot_data, aes(x = scDblFinder_status, y = library_size)) + geom_violin(fill = "red") + scale_y_log10()


```

```{r}
is.mito <- grep(rowData(sce_scdblfinder)$Symbol, pattern = "^MT-")

is.mito <- grep("^MT-", 
                rowData(sce_scdblfinder)$Symbol,
                value = FALSE)

sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder, 
                             subsets = list(MT = is.mito))

plot_data <- tibble(library_size = colSums(assay(sce_scdblfinder, "counts")), scDblFinder_status = sce_scdblfinder$scDblFinder.class, mito.percentage = sce_scdblfinder$subsets_MT_percent)
  

ggplot(plot_data, aes(x = scDblFinder_status, y = mito.percentage)) + geom_violin(fill = "blue") + scale_y_log10()
  

```

```{r}
sce_scdblfinder <- logNormCounts(sce_scdblfinder)

sce_scdblfinder <- scater::runPCA(sce_scdblfinder, 
                      ncomponents = 50)

reducedDim(sce_scdblfinder)

plot_data <- tibble(library_size = colSums(assay(sce_scdblfinder, "counts")), 
                    scDblFinder_status = sce_scdblfinder$scDblFinder.class, 
                    mito.percentage = sce_scdblfinder$subsets_MT_percent,
                    PC1 = reducedDim(sce_scdblfinder, type = "PCA")[,1],
                    PC2 = reducedDim(sce_scdblfinder, type = "PCA")[,2])

ggplot(plot_data, 
       aes(x = PC1, y = PC2)) + geom_point(aes(colour = scDblFinder_status))
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

set.seed(1)
sce_decontX <- celda::decontX(sce_filtered_scdblfinder,
                              assayName = "counts")


```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
umap <- plotDimReduceCluster(sce_decontX$decontX_clusters, 
                     dim1 = reducedDim(sce_decontX)[,1],
                     dim2 = reducedDim(sce_decontX)[,2])

umap
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)


```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}
marker_genes <- c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1")

marker_positions <- grep(paste(marker_genes,collapse="|"),
                         rowData(sce_decontX)$Symbol,
                         value = FALSE)

marker_names <- rowData(sce_decontX)$ID[marker_positions]




rowData(sce_decontX)
rownames(sce_decontX) <-
  uniquifyFeatureNames(rowData(sce_decontX)$ID, rowData(sce_decontX)$Symbol)

plot_cont <- plotDimReduceFeature(
  decontXcounts(sce_decontX), # logcounts will show the contaminated data
  features = c(
    "CD3D",
    "CD3E",
    "GNLY",
    "LYZ",
    "S100A8",
    "S100A9",
    "CD79A",
    "CD79B",
    "MS4A1", 
    "PPBP", 
    "FCER1A", "CST3"),
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
  
)
  
  
plot_cont
  
  

```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
markers <- list("T cells" = c("CD3D", "CD3E"),
                "monocytes" = c("LYZ", "S100A8", "S100A9"),
                "B cells" = c("CD79A", "CD79B", "MS4A1"),
                "NK cells" = "GNLY",
                "megakaryocytes" = "PPBP",
                "dendritic cells" = c("FCER1A", "CST3"))
  


cellTypeMappings <- list(
  Tcells = 1,
  monocytes = 2,
  Bcells = 3,
  NKcells = 7,
  megakaryocyte = 8
  
)

plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts"))
```


- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r, fig.height=10, fig.width=10}


sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts", name = "logcounts")
sce_decontX <- logNormCounts(sce_decontX, assay.type = "decontXcounts", name = "logdecontXcounts")

assays(sce_decontX)

plotDecontXMarkerExpression(sce_decontX,
                            markers = markers,
                            groupClusters = cellTypeMappings,
                            assayName = c("logcounts", "logdecontXcounts"),
                            log1p = F)
  
  
```
