---
title: "Example code for dimensionality reduction and clustering in R"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(umap)
library(Rtsne)
library(dendextend)
library(dbscan)
```

# Exercise

## Setup

- Import the `iris` data set.

```{r}
data("iris")

```

- Separate the matrix of measurements in a new object named `iris_features`.

```{r}
iris_features <- iris %>% select(-Species)

head(iris_features)
```

# Exercise

## Apply Principal Components Analysis (PCA)

The `prcomp()` function allows you to standardise the data as part of the principal components analysis itself.

- Apply PCA while centering and scaling the matrix of features.

```{r}
pca_iris <- prcomp(iris_features, center = TRUE, scale. = TRUE)
pca_iris


```

- Examine the PCA output.
  Display the loading of each feature on each principal component.

```{r}
pca_iris$x

summary(pca_iris)

str(pca_iris)
```


- Use the return value of the PCA to create a `data.frame` called `pca_iris_dataframe` that contains the coordinates projected on principal components.

```{r}
pca_iris_dataframe <- pca_iris$x
pca_iris_dataframe <- as.data.frame(pca_iris_dataframe)
head(pca_iris_dataframe)
```

- Visualise the PCA projection using `ggplot2::geom_point()`.

```{r}
ggplot(pca_iris_dataframe) +
  geom_point(aes(x = PC1, y = PC2))
  
  
```

### Bonus point

- Color data points according to their species

- Store the PCA plot as an object named `pca_iris_species`.

```{r}

iris_species <- iris$Species

pca_iris_df_species <- cbind(pca_iris_dataframe, iris_species)

head(pca_iris_df_species)
```

```{r}
pca_iris_species_12 <- ggplot(pca_iris_df_species) +
  geom_point(aes(x = PC1, y = PC2, colour = iris_species)) +
  ggtitle("PCA") +
  theme(plot.title = element_text(hjust = 0.5))

pca_iris_species_12

pca_iris_species_34 <- ggplot(pca_iris_df_species) +
  geom_point(aes(x = PC3, y = PC4, colour = iris_species))

pca_iris_species_34
```

# Exercise

## Variable loading

- Color a scatter plot of PC1 and PC2 by the value of the variable most strongly associated with the first principal component.

What do you observe?

```{r}

pca_iris$rotation

pca_iris_col <- ggplot(pca_iris_df_species) +
  geom_point(aes(x = PC1, y = PC2, colour = iris$Petal.Length)) #can also take out petal length and cbind as done with species above

pca_iris_col
```

``
> 

## Variance explained

- Compute the variance explained by principal components, using information present in the return value of the `prcomp()` function.

```{r}
explained_variance_ratio <- pca_iris$sdev^2 / sum(pca_iris$sdev^2)
explained_variance_ratio


pca_ratio_summary <- summary(pca_iris)
```

- Visualise the variance explained by each principal component using `ggplot2::geom_col()`.

```{r}
variance_dataframe <- data.frame(explained_variance_ratio)

variance_dataframe$PC <- c("PC1", "PC2", "PC3", "PC4")
head(variance_dataframe)
```

```{r}
ggplot(variance_dataframe, aes(x = PC, y = explained_variance_ratio)) + 
  geom_col()

```

# Exercise

## UMAP

- Apply UMAP on the output of the PCA.

```{r}
set.seed(1) # Set a seed for reproducible results
umap_iris <- umap(pca_iris_dataframe)

umap_iris2 <- umap(iris[,1:4])

```

- Inspect the UMAP output.

```{r}
head(umap_iris$layout)
```

- Visualise the UMAP projection using `ggplot2::geom_point()`.

```{r}
umap_iris_dataframe <- as.data.frame(umap_iris$layout)

head(umap_iris_dataframe)

umap_iris2_df <- as.data.frame(umap_iris2$layout)
```

```{r}
ggplot(umap_iris_dataframe, aes(x = V1, y = V2)) + 
  geom_point()

ggplot(umap_iris2_df, aes(x = V1, y = V2)) + 
  geom_point()
  
```

### Bonus point

- Color data points according to their class label.

- Store the UMAP plot as an object named `umap_iris_species`.

```{r}
head(umap_iris_dataframe)
```

```{r}
umap_iris_species <- ggplot(umap_iris_dataframe, aes(x = V1, y = V2, colour = iris$Species)) + 
  geom_point() +
  ggtitle("UMAP") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
umap_iris_species

umap_iris_species2 <- ggplot(umap_iris2_df, aes(x = V1, y = V2, colour = iris$Species)) + 
  geom_point() +
  ggtitle("UMAP2") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
umap_iris_species2
```

# Exercise

## t-SNE

- Apply t-SNE and inspect the output.

```{r}
set.seed(1) # Set a seed for reproducible results
tsne_iris <- Rtsne(pca_iris$x, dims = 2, initial_dims = 4, check_duplicates = FALSE, pca = FALSE)

str(tsne_iris)

tsne_iris$Y
```

- Use the return value of the t-SNE to create a `data.frame` called `tsne_iris_dataframe` that contains the coordinates.

```{r}
tsne_iris_dataframe <- as.data.frame(tsne_iris$Y)

head(tsne_iris_dataframe)
```

- Visualise the t-SNE projection.

```{r}
ggplot(tsne_iris_dataframe, aes(x = V1, y = V2)) + geom_point()
  
  
```

### Bonus points

- Color data points according to their class label.

- Store the t-SNE plot as an object named `tsne_iris_species`.

```{r}

head(tsne_iris_dataframe)
```

```{r}
tsne_iris_species <- 
  ggplot(tsne_iris_dataframe, aes(x = V1, y = V2, colour = iris$Species)) +
  geom_point() +
  ggtitle("t-SNE") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
tsne_iris_species
```

- Combine PCA, UMAP and t-SNE plots in a single figure.

```{r, fig.height=2, fig.width=6}
cowplot::plot_grid(pca_iris_species_12, umap_iris_species , tsne_iris_species, nrow = 1)
  
```

# Exercise

## Hierarchical clustering

- Perform hierarchical clustering on the `iris_features` data set,
  using the `euclidean` distance and method `ward.D2`.
  Use the functions `dist()` and `hclust()`.

```{r}
dist_iris <- dist(iris_features, method = "euclidean")
hclust_iris_ward <- hclust(dist_iris, method = "ward.D2")

```

- Plot the clustering tree.
  Use the function `plot()`.

```{r}
plot(hclust_iris_ward)
```

How many clusters would you call from a visual inspection of the tree?

> Answer: 3 or 6
> 
> 

- **Bonus point:** Color leaves by known species (use `dendextend`).

```{r}
iris_hclust_dend <- as.dendrogram(hclust_iris_ward)
labels_colors(iris_hclust_dend) <- as.numeric(iris$Species)
plot(iris_hclust_dend)
```

- Cut the tree in 3 clusters and extract the cluster label for each flower.
  Use the function `cutree()`.

```{r}
iris_hclust_ward_labels <- cutree(hclust_iris_ward, k = 3)
iris_hclust_ward_labels
```

- Repeat clustering using 3 other agglomeration methods:

  + `complete`
  + `average`
  + `single`

```{r}
# complete
hclust_iris_complete <- hclust(dist_iris, method = "complete")

iris_hclust_complete_labels <- cutree(hclust_iris_complete, k = 3)
iris_hclust_complete_labels
```

```{r}
# average
hclust_iris_average <- hclust(dist_iris, method = "average")
iris_hclust_average_labels <- cutree(hclust_iris_average, k=3)
iris_hclust_average_labels
```

```{r}
# single
hclust_iris_single <- hclust(dist_iris, method = "single")
iris_hclust_single_labels <- cutree(hclust_iris_single, k = 3)
iris_hclust_single_labels
```

- Compare clustering results on scatter plots of the data.

```{r}
iris_clusters_dataframe <- iris
iris_clusters_dataframe$hclust_average <- as.factor(iris_hclust_average_labels)
iris_clusters_dataframe$hclust_complete <- as.factor(iris_hclust_average_labels)
iris_clusters_dataframe$hclust_single <- as.factor(iris_hclust_single_labels)
iris_clusters_dataframe$hclust_ward <- as.factor(iris_hclust_ward_labels)
```

```{r, fig.height=4, fig.width=5}
plot_average <- ggplot(iris_clusters_dataframe, aes(x = Petal.Width, y = Petal.Length, colour = hclust_average)) + geom_point() +
  ggtitle("Average") +
  theme(plot.title = element_text(hjust = 0.5))

plot_complete <- ggplot(iris_clusters_dataframe, aes(x = Petal.Width, y = Petal.Length, colour = hclust_complete)) + geom_point() +
  ggtitle("Complete") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
plot_single <- ggplot(iris_clusters_dataframe, aes(x = Petal.Width, y = Petal.Length, colour = hclust_single)) + geom_point() +
  ggtitle("Single") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
plot_ward <- ggplot(iris_clusters_dataframe, aes(x = Petal.Width, y = Petal.Length, colour = hclust_ward)) + geom_point() +
  ggtitle("Ward") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
cowplot::plot_grid(plot_average, plot_complete, plot_single, plot_ward, nrow = 2)
  

```

# Exercise

## dbscan

- Apply `dbscan` to the `iris_features` data set.

```{r}
dbscan_iris <- dbscan(iris_features, eps = 0.7)
dbscan_iris
```

- Visualise the `dbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$dbscan <- as.factor(dbscan_iris$cluster)
head(iris_clusters_dataframe)
```

```{r}
dbscan_plot <- ggplot(iris_clusters_dataframe, aes(x = Petal.Width, y = Petal.Length, colour = dbscan)) + geom_point() +
  ggtitle("DBScan") +
  theme(plot.title = element_text(hjust = 0.5))
  
dbscan_plot
```

## hdbscan

- Apply `hdbscan` to the `iris_features` data set.

```{r}
hdbscan_iris <- hdbscan(iris_features, minPts = 3)
hdbscan_iris
```

- Visualise the `hdbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$hdbscan <- as.factor(hdbscan_iris$cluster)
head(iris_clusters_dataframe)
```

```{r}
hdbscan_plot <- ggplot(iris_clusters_dataframe, aes(x = Petal.Width, y = Petal.Length, colour = hdbscan)) + 
  geom_point() +
  ggtitle("DBScan") +
  theme(plot.title = element_text(hjust = 0.5))
  
  
hdbscan_plot
```

## Bonus point

- Combine the plots of `dbscan` and `hdbscan` into a single plot.

```{r, fig.height=3, fig.width=6}
cowplot::plot_grid(dbscan_plot, hdbscan_plot)
  

```

# Exercise

## K-means clustering

- Apply $K$-means clustering with $K$ set to 3 clusters.

```{r}
set.seed(1) # Set a seed for reproducible results
kmeans_iris <- kmeans(   )
kmeans_iris
```

- Inspect the output.

```{r}

```

- Extract the cluster labels.

```{r}

```

- Extract the coordinates of the cluster centers.

```{r}

```

- Construct a data frame that combines the `iris` dataset and the cluster label.

```{r}
iris_labelled <- iris
iris_labelled$Kmeans <- as.factor(   )
head(iris_labelled)
```

- Plot the data set as a scatter plot.

  + Color by cluster label.

```{r}
ggplot(iris_labelled,    ) +
  
  
```

### Bonus point

- Add cluster centers as points in the plot.

```{r}
iris_means_centers <- as.data.frame(   )
iris_means_centers$Kmeans <- as.factor(   )
head(iris_means_centers)
```


```{r}
ggplot(iris_labelled,    ) +
  
  
  
```

# Exercise

## Cross-tabulation with ground truth

- Cross-tabulate cluster labels with known labels.

```{r}
table(   )
```

How many observations are mis-classified by $K$-means clustering?

> Answer:
> 
> 
> 
> 
> 

## Elbow plot

- Plot the "total within-cluster sum of squares" for K ranging from 2 to 10.

```{r}

```

```{r}
get_mean_totss_for_k <- function(k, data) {
  kmeans_out <- kmeans(data, k)
  return(kmeans_out$tot.withinss)
}
k_range <- 2:10
kmean_totwithinss <- vapply(   )
kmean_totwithinss
```

```{r}
kmean_totwithinss_dataframe <- data.frame(
  K = ,
  totss = 
)
head(kmean_totwithinss_dataframe)
```

```{r}
ggplot(kmean_totwithinss_dataframe,    ) +
  
  
  
```

Do you agree that 3 is the optimal number of clusters for this data set?

> Answer:
> 
> 
> 
> 

