---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}

data("diamonds")
ggplot(diamonds)
```

> Answer: grey box; nothing is specified

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer: you have plotted the x- and y-axis of the values you have assigned to it, but you have not told it how to display it (geom)

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(shape = 21)
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) + 
  geom_point(shape = 21)
  
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) + 
  geom_point(shape = 21) + geom_smooth()


```

- Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(aes(colour = cut), shape = 21) + geom_smooth()


```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer: ggplot takes the information specified in the ggplot argument if nothing is specified in the specific sub-argument

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
#load dataset ChickWeight
data("ChickWeight")
colnames(ChickWeight)

ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point()

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = lm)
```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer: diet 3

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Diet, nrow = 1)

#ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  #geom_point() +
  #geom_smooth(method = lm) +
  #facet_wrap(~ Chick, nrow = 10)

```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~ Diet, nrow = 1)


```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}

msleep <- ggplot2::msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x = order)) + geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) + geom_bar() + theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1))

```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) + 
  geom_bar() + theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1)) + 
  theme(axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(x = "Order", y = "Count")

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) + 
  geom_bar(aes(fill = genus)) + 
  theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1)) + 
  theme(axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(x = "Order", y = "Count")

```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) + 
  geom_bar(aes(fill = genus)) + 
  theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1)) + 
  theme(axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(x = "Order", y = "Count") +
  theme(legend.key.size = unit(2, "mm"))


```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x = order)) + 
  geom_bar(aes(fill = genus)) + 
  theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1)) + 
  theme(axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(x = "Order", y = "Count") +
  theme(legend.key.size = unit(2, "mm")) +
  guides(fill = guide_legend(ncol = 3))



```

- Add a contour of thin black lines to the bars.

```{r}


ggplot(msleep, aes(x = order)) + 
  geom_bar(aes(fill = genus), colour = "black", linewidth = 0.1) + 
  theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1)) + 
  theme(axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(x = "Order", y = "Count") +
  theme(legend.key.size = unit(2, "mm")) +
  guides(fill = guide_legend(ncol = 3))



```

- Assign that last plot to an object called `obds_msleep`.

```{r}

obds_msleep <- ggplot(msleep, aes(x = order)) + 
  geom_bar(aes(fill = genus), colour = "black", linewidth = 0.1) + 
  theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1)) + 
  theme(axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(x = "Order", y = "Count") +
  theme(legend.key.size = unit(2, "mm")) +
  guides(fill = guide_legend(ncol = 3)) #fill is what is in your columns, so change to "colour =" or other

```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=8, fig.width=12}
library(cowplot)



first_row <- cowplot::plot_grid(obds_diamonds, obds_chickweight, labels = c("A", "B"))


second_row <- cowplot::plot_grid(obds_msleep, labels = "C")


combined_plot <- cowplot::plot_grid(first_row, second_row, ncol = 1)
  
first_row #define first row
second_row #define second row
combined_plot #combined first and second row


superplot <- cowplot::plot_grid(obds_diamonds, obds_chickweight, obds_msleep, labels = c("A", "B", "C"), nrow = 2) #makes it as a grid so row 1, column 1 = plot 1, r1/c2 = plot 2, r2/c1 = plot 3
superplot
```

Using `patchwork`.

```{r, fig.height=3, fig.width=12}
library(patchwork)

obds_chickweight + obds_diamonds + obds_msleep #next to each other

patch_plot <- (obds_chickweight + obds_diamonds) / obds_msleep #plots1-2 on row 1, plot 3 on row 2
```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}

library(gridExtra)

grid_plot <- gridExtra::grid.arrange(obds_chickweight, obds_diamonds, obds_msleep, ncol = 2)

```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}

ggsave("combined_plot.pdf", combined_plot, width = 12, height = 10, units = "in")

ggsave("superplot.pdf", superplot, width = 12, height = 8, units = "in")

ggsave("patch_plot.pdf", patch_plot, width = 12, height = 8, units = "in")

```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}
mpg <- ggplot2::mpg

ggplot(mpg, aes(x = ))

```
